---
title: "NBA 2025-26 Salary vs. Performance"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    theme: default
    source_code: embed
---

```{r setup, include=FALSE}
library(flexdashboard)
library(tidyverse)
library(plotly)
library(scales)
library(readr)
```

```{r, include=FALSE}

stats    <- read.csv("data/nba-2026-stats.csv")
salaries <- read.csv("data/nba-2026-salaries.csv")

glimpse(stats)
glimpse(salaries)

```

```{r, include=FALSE}

salaries_clean <- salaries %>%
  select(Player, Tm, salary_raw = X2025.26) %>%
  mutate(
    salary = as.numeric(gsub("[$, ]", "", salary_raw))
  ) %>%
  select(Player, Tm, salary) %>%
  filter(!is.na(salary)) %>%
  group_by(Player) %>%
  slice_tail(n = 1) %>%
  ungroup()

stats_clean <- stats %>%
  filter(!is.na(PTS)) %>%
  group_by(Player) %>%
  filter(if (any(Team %in% c("TOT", "2TM", "3TM", "4TM", "5TM"))) 
           Team %in% c("TOT", "2TM", "3TM", "4TM", "5TM") 
         else 
           row_number() == 1) %>%
  slice(1) %>% 
  ungroup()

nba <- stats_clean %>%
  full_join(salaries_clean, by = "Player") %>%
  mutate(
    Team = coalesce(Team, Tm),  # fill Team from contracts if missing from stats
    G    = replace_na(G, 0),
    PTS  = replace_na(PTS, 0),
    AST  = replace_na(AST, 0),
    TRB  = replace_na(TRB, 0),
    STL  = replace_na(STL, 0),
    BLK  = replace_na(BLK, 0),
    FGA  = replace_na(FGA, 0),
    FG   = replace_na(FG, 0),
    FTA  = replace_na(FTA, 0),
    FT   = replace_na(FT, 0),
    TOV  = replace_na(TOV, 0)
  )

glimpse(nba)
```

```{r, include=FALSE}

nba <- nba %>%
  mutate(
    salary_m         = salary / 1e6,
    pts_per_million  = PTS / salary_m,
    value_score = (PTS + AST + TRB) / pmax(salary_m, 1) * (G / 82),
    efficiency       = (PTS + TRB + AST + STL + BLK) / 
                       (FGA - FG + FTA * 0.44 - FT + TOV),
   salary_tier = case_when(
  salary_m >= 40  ~ "Max ($40M+)",
  salary_m >= 25  ~ "All-Star ($25-40M)",
  salary_m >= 10  ~ "Starter ($10-25M)",
  salary_m >= 2   ~ "Rotation ($2-10M)",
  TRUE            ~ "Minimum (< $2M)"
),
salary_tier = factor(salary_tier,
                     levels = c("Max ($40M+)", "All-Star ($25-40M)",
                                "Starter ($10-25M)", "Rotation ($2-10M)",
                                "Minimum (< $2M)")))

glimpse(nba)

```
The Big Picture
=====================================

Row {data-height=140}
-------------------------------------

### League Median Salary

```{r}

valueBox(dollar(median(nba$salary, na.rm = TRUE), accuracy = 1),
         caption = "League Median Salary",
         color = "#1a1a2e")

```

### % Overpaid Players

```{r}


overpaid <- nba %>% 
  filter(salary_m >= 10) %>%
  summarise(pct = mean(PTS < mean(nba$PTS, na.rm = TRUE), na.rm = TRUE))
valueBox(percent(overpaid$pct, accuracy = 0.1),
         caption = "Of $10M+ Players Score Below League Avg PPG",
         color = "#1a1a2e")

```

### Best Value Player

```{r}

best <- nba %>% filter(G >= 30, salary_m >= 5) %>% 
        slice_max(value_score, n = 1)
valueBox(paste0(best$Player, " (", best$Team, ")"),
         caption = paste0("$", round(best$salary_m, 1), "M | Value Score: ", 
                          round(best$value_score, 1)),
         color = "#1a1a2e")

```


### Biggest Bust
```{r}

bust <- nba %>% filter(G >= 30, salary_m >= 25) %>%
        slice_min(efficiency, n = 1)
valueBox(paste0(bust$Player, " (", bust$Team, ")"),
         caption = paste0("$", round(bust$salary_m, 1), "M | Efficiency: ", 
                          round(bust$efficiency, 2)),
         color = "#1a1a2e")

```

Row {data-height=430}
-------------------------------------

### Salary vs. Points Per Game

```{r}

plot_ly(nba %>% filter(G >= 20),
        x      = ~salary_m,
        y      = ~PTS,
        color  = ~salary_tier,
        colors = c("#ff6b35", "#00e5a0", "#7b5ea7", "#4ecdc4", "#ffe66d"),
        type   = "scatter",
        mode   = "markers",
        marker = list(size = 7, opacity = 0.7),
        text   = ~paste0(Player, "<br>",
                         "Salary: $", round(salary_m, 1), "M<br>",
                         "PPG: ", PTS, "<br>",
                         "Team: ", Team),
        hovertemplate = "%{text}<extra></extra>") %>%
  layout(
    xaxis = list(title = "2025-26 Salary ($M)", tickprefix = "$"),
    yaxis = list(title = "Points Per Game"),
    legend = list(orientation = "h", y = -0.2)
  )

```


Row {data-height=430}
-------------------------------------
### Value Score by Team

```{r}

team_summary <- nba %>%
  filter(salary_m >= 1) %>% 
  group_by(Team) %>%
  summarise(
    avg_value     = mean(value_score, na.rm = TRUE),
    avg_salary    = mean(salary_m, na.rm = TRUE),
    avg_pts       = mean(PTS, na.rm = TRUE),
    total_payroll = sum(salary_m, na.rm = TRUE),
    players       = n(),
    .groups = "drop"
  ) %>%
  filter(!Team %in% c("2TM", "3TM", "4TM", "5TM")) %>%
  arrange(desc(avg_value))

plot_ly(team_summary,
        x = ~reorder(Team, avg_value),
        y = ~avg_value,
        type = "bar",
        marker = list(color = ~avg_value,
                      colorscale = list(c(0, "#ff6b35"), c(0.5, "#7b5ea7"), c(1, "#00e5a0")),
                      showscale = FALSE),
        text = ~paste0(Team, "<br>",
               "Avg Value Score: ", round(avg_value, 1), "<br>",
               "Avg Salary: $", round(avg_salary, 1), "M<br>",
               "Total Payroll: $", round(total_payroll, 1), "M<br>",
               "Players included: ", players),
        hovertemplate = "%{text}<extra></extra>") %>%
  layout(
    xaxis = list(title = "Team", tickangle = -45),
    yaxis = list(title = "Avg Value Score"),
    bargap = 0.2
  )

```


Row {data-height=120}
-------------------------------------

###
**Efficiency Score** — (PTS + TRB + AST + STL + BLK) / (missed FG + 0.44 × missed FT + TOV). Higher is better. &nbsp;|&nbsp; **Value Score** — (PTS + AST + TRB) per $1M salary.

**Data Notes** — Players on Exhibit 10 / two-way contracts (salary < $1M) excluded from team value analysis. Traded players use combined season totals. Players with contracts but zero games played (e.g. Anthony Davis, Trae Young on WAS) are included with $0 value score — their salaries count against team payroll. Source: Basketball Reference 2025-26.

Value Kings & Busts
=====================================

Row {data-height=500}
-------------------------------------

### Top 15 Value Players

```{r}

top15 <- nba %>%
  filter(G >= 30, salary_m >= 2) %>%
  slice_max(value_score, n = 15) %>%
  arrange(value_score)

plot_ly(top15,
        x    = ~value_score,
        y    = ~reorder(Player, value_score),
        type = "bar",
        orientation = "h",
        marker = list(color = "#00e5a0"),
        text = ~paste0(Player, "<br>",
                       "Salary: $", round(salary_m, 1), "M<br>",
                       "Value Score: ", round(value_score, 1), "<br>",
                       "Team: ", Team),
        hovertemplate = "%{text}<extra></extra>") %>%
  layout(
    xaxis = list(title = "Value Score (PTS+AST+TRB per $1M)"),
    yaxis = list(title = ""),
    margin = list(l = 150)
  )

```


### Bottom 15 — Most Overpaid

```{r}

bot15 <- nba %>%
  filter(G >= 30, salary_m >= 15) %>%
  slice_min(value_score, n = 15) %>%
  arrange(desc(value_score))

plot_ly(bot15,
        x    = ~value_score,
        y    = ~reorder(Player, -value_score),
        type = "bar",
        orientation = "h",
        marker = list(color = "#ff6b35"),
        text = ~paste0(Player, "<br>",
                       "Salary: $", round(salary_m, 1), "M<br>",
                       "Value Score: ", round(value_score, 1), "<br>",
                       "Team: ", Team),
        hovertemplate = "%{text}<extra></extra>") %>%
  layout(
    xaxis = list(title = "Value Score (PTS+AST+TRB per $1M)"),
    yaxis = list(title = ""),
    margin = list(l = 150)
  )

```


Row {data-height=500}
-------------------------------------

### Value Score vs. Salary

```{r}

plot_ly(nba %>% filter(G >= 20, salary_m >= 2),
        x     = ~salary_m,
        y     = ~value_score,
        color = ~salary_tier,
        colors = c("#ff6b35", "#00e5a0", "#7b5ea7", "#4ecdc4", "#ffe66d"),
        type  = "scatter",
        mode  = "markers",
        marker = list(size = 7, opacity = 0.7),
        text  = ~paste0(Player, "<br>",
                        "Salary: $", round(salary_m, 1), "M<br>",
                        "Value Score: ", round(value_score, 1), "<br>",
                        "PPG: ", PTS, "<br>",
                        "Games: ", G, "<br>",
                        "Team: ", Team),
        hovertemplate = "%{text}<extra></extra>") %>%
  layout(
    xaxis = list(title = "2025-26 Salary ($M)", tickprefix = "$"),
    yaxis = list(title = "Value Score per $1M (games-adjusted)",
             range = c(0, 10)),
    legend = list(orientation = "h", y = -0.2),
    shapes = list(list(
      type = "line", x0 = 0, x1 = 1, xref = "paper",
      y0 = median(nba$value_score, na.rm = TRUE),
      y1 = median(nba$value_score, na.rm = TRUE),
      line = list(color = "#ff6b35", dash = "dash", width = 1.5)
    )),
    annotations = list(list(
      x = 0.01, xref = "paper",
      y = median(nba$value_score, na.rm = TRUE),
      text = "League Median Value Score",
      showarrow = FALSE,
      font = list(color = "#ff6b35", size = 10),
      yshift = 10
    ))
  )

```

Row {data-height=120}
-------------------------------------

###
**Reading These Charts** — Value Score = (PTS + AST + TRB) per $1M salary, scaled by games played rate (G/82). A player earning $5M with 20 pts/7 reb/5 ast over 75 games scores far higher than a $50M player with similar stats over 40 games. Minimum salary floor of $2M applied to all charts to remove two-way and exhibit 10 contracts.

**Top 15 filter** — Players must have appeared in 30+ games and earn $2M+. **Bust filter** — Players must have appeared in 30+ games and earn $15M+. Players below median value score on max contracts represent the clearest cases of salary inefficiency in the league. Note that injury-limited players (e.g. Darius Garland, 58 games) are penalized by the games rate component — this is intentional, as availability is a core component of player value. Source: Basketball Reference 2025-26.

Salary Tiers
=====================================

Row {data-height=500}
-------------------------------------

### PPG Distribution by Salary Tier

```{r}

plot_ly(nba %>% filter(G >= 20, salary_m >= 2, !is.na(salary_tier)),
        x      = ~salary_tier,
        y      = ~PTS,
        color  = ~salary_tier,
        colors = c("#ff6b35", "#00e5a0", "#7b5ea7", "#4ecdc4", "#ffe66d"),
        type   = "box",
        hovertemplate = "Player: %{text}<br>PPG: %{y}<extra></extra>",
        text   = ~Player) %>%
  layout(
    showlegend = FALSE,
    xaxis = list(title = "Salary Tier"),
    yaxis = list(title = "Points Per Game")
  )

```


### Position Breakdown by Salary Tier

```{r}

pos_tier <- nba %>%
  filter(!is.na(salary_tier), salary_m >= 2,
         Pos %in% c("PG", "SG", "SF", "PF", "C")) %>%
  count(salary_tier, Pos) %>%
  group_by(salary_tier) %>%
  mutate(pct = n / sum(n))

plot_ly(pos_tier,
        x      = ~salary_tier,
        y      = ~pct,
        color  = ~Pos,
        colors = c("#00e5a0", "#7b5ea7", "#ff6b35", "#4ecdc4", "#ffe66d"),
        type   = "bar",
        text   = ~paste0(Pos, ": ", n, " players (", percent(pct, accuracy = 0.1), ")"),
        hovertemplate = "%{text}<extra></extra>") %>%
  layout(
    barmode = "stack",
    xaxis = list(title = "Salary Tier"),
    yaxis = list(title = "% of Players", tickformat = ".0%"),
    legend = list(orientation = "h", y = -0.2)
  )

```


Row {data-height=500}
-------------------------------------

### Punching Above Their Tier

```{r}

tier_medians <- nba |>
  filter(!is.na(salary_tier), salary_m >= 2, G >= 20) |>
  group_by(salary_tier) |>
  summarise(tier_median = median(value_score, na.rm = TRUE), .groups = "drop")

punchers <- nba |>
  filter(!is.na(salary_tier), salary_m >= 2, G >= 20) |>
  left_join(tier_medians, by = "salary_tier") |>
  mutate(above_tier = value_score - tier_median) |>
  group_by(salary_tier) |>
  slice_max(above_tier, n = 5) |>
  ungroup() |>
  arrange(salary_tier, desc(above_tier))

plot_ly(punchers,
        x    = ~above_tier,
        y    = ~reorder(Player, above_tier),
        color = ~salary_tier,
        colors = c("#ff6b35", "#00e5a0", "#7b5ea7", "#4ecdc4"),
        type = "bar",
        orientation = "h",
        text = ~paste0(Player, " (", Team, ")<br>",
                       "Salary: $", round(salary_m, 1), "M<br>",
                       "Value Score: ", round(value_score, 1), "<br>",
                       "Above tier median: +", round(above_tier, 1)),
        hovertemplate = "%{text}<extra></extra>") |>
  layout(
    xaxis = list(title = "Value Score Above Tier Median"),
    yaxis = list(title = ""),
    legend = list(orientation = "h", y = -0.15),
    margin = list(l = 150)
  )

```

Position Analysis
=====================================

Row {data-height=500}
-------------------------------------

### Avg Stats by Position

```{r}
pos_stats <- nba |>
  filter(G >= 20, salary_m >= 2,
         Pos %in% c("PG", "SG", "SF", "PF", "C")) |>
  group_by(Pos) |>
  summarise(
    PPG  = round(mean(PTS, na.rm = TRUE), 1),
    APG  = round(mean(AST, na.rm = TRUE), 1),
    RPG  = round(mean(TRB, na.rm = TRUE), 1),
    SPG  = round(mean(STL, na.rm = TRUE), 1),
    BPG  = round(mean(BLK, na.rm = TRUE), 1),
    .groups = "drop"
  ) |>
  pivot_longer(-Pos, names_to = "stat", values_to = "value")

plot_ly(pos_stats,
        x      = ~Pos,
        y      = ~value,
        color  = ~stat,
        colors = c("#00e5a0", "#7b5ea7", "#ff6b35", "#4ecdc4", "#ffe66d"),
        type   = "bar",
        barmode = "group",
        text   = ~paste0(stat, ": ", value),
        hovertemplate = "%{text}<extra></extra>") |>
  layout(
    barmode = "group",
    xaxis = list(title = "Position", 
             categoryorder = "array",
             categoryarray = c("PG", "SG", "SF", "PF", "C")),
    yaxis = list(title = "Per Game Average"),
    legend = list(orientation = "h", y = -0.2)
  )
```


### Salary vs. Efficiency by Position

```{r}
plot_ly(nba |> filter(G >= 20, salary_m >= 2,
                      Pos %in% c("PG", "SG", "SF", "PF", "C")),
        x      = ~salary_m,
        y      = ~efficiency,
        color  = ~Pos,
        colors = c("#00e5a0", "#7b5ea7", "#ff6b35", "#4ecdc4", "#ffe66d"),
        type   = "scatter",
        mode   = "markers",
        marker = list(size = 7, opacity = 0.7),
        text   = ~paste0(Player, "<br>",
                         "Pos: ", Pos, "<br>",
                         "Salary: $", round(salary_m, 1), "M<br>",
                         "Efficiency: ", round(efficiency, 2)),
        hovertemplate = "%{text}<extra></extra>") |>
  layout(
    xaxis = list(title = "Salary ($M)", tickprefix = "$"),
    yaxis = list(title = "Efficiency Score"),
    legend = list(orientation = "h", y = -0.2)
  )
```



Row {data-height=500}
-------------------------------------

### Position Value Score Comparison

```{r}

pos_value <- nba |>
  filter(G >= 20, salary_m >= 2,
         Pos %in% c("PG", "SG", "SF", "PF", "C")) |>
  group_by(Pos) |>
  summarise(
    avg_value    = mean(value_score, na.rm = TRUE),
    median_value = median(value_score, na.rm = TRUE),
    avg_salary   = mean(salary_m, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(Pos = factor(Pos, levels = c("PG", "SG", "SF", "PF", "C")))

plot_ly(pos_value) |>
  add_trace(x = ~Pos, y = ~avg_value,
            type = "bar", name = "Avg Value Score",
            marker = list(color = "#00e5a0"),
            hovertemplate = "Avg Value Score: %{y:.2f}<extra></extra>") |>
  add_trace(x = ~Pos, y = ~median_value,
            type = "bar", name = "Median Value Score",
            marker = list(color = "#7b5ea7"),
            hovertemplate = "Median Value Score: %{y:.2f}<extra></extra>") |>
  layout(
    barmode = "group",
    xaxis = list(title = "Position",
                 categoryorder = "array",
                 categoryarray = c("PG", "SG", "SF", "PF", "C")),
    yaxis = list(title = "Value Score"),
    legend = list(orientation = "h", y = -0.2)
  )

```

Star Player Spotlight
=====================================

Row {data-height=500}
-------------------------------------

### Salary vs. Value Score — Elite Players

```{r}

elite <- nba |> filter(PTS >= 20, G >= 30, salary_m >= 2)

plot_ly(elite,
        x      = ~salary_m,
        y      = ~value_score,
        color  = ~Pos,
        colors = c("#00e5a0", "#7b5ea7", "#ff6b35", "#4ecdc4", "#ffe66d"),
        type   = "scatter",
        mode   = "markers",
        marker = list(size = 10, opacity = 0.8),
        customdata = ~paste0("<b>", Player, "</b><br>",
                     "Salary: $", round(salary_m, 1), "M<br>",
                     "Value Score: ", round(value_score, 1), "<br>",
                     "PPG: ", PTS, "<br>",
                     "Team: ", Team),
        hovertemplate = "%{customdata}<extra></extra>") |>
  layout(
    xaxis = list(title = "2025-26 Salary ($M)", tickprefix = "$"),
    yaxis = list(title = "Value Score"),
    legend = list(orientation = "h", y = -0.2),
    shapes = list(list(
      type = "line", x0 = 0, x1 = 1, xref = "paper",
      y0 = median(elite$value_score, na.rm = TRUE),
      y1 = median(elite$value_score, na.rm = TRUE),
      line = list(color = "#ff6b35", dash = "dash", width = 1.5)
    )),
    annotations = list(list(
      x = 0.01, xref = "paper",
      y = median(elite$value_score, na.rm = TRUE),
      text = "Elite Median Value Score",
      showarrow = FALSE,
      font = list(color = "#ff6b35", size = 10),
      yshift = 10
    ))
  )

```


Row {data-height=500}
-------------------------------------

### Elite Player Rankings

```{r}

elite_table <- nba |>
  filter(PTS >= 20, G >= 30, salary_m >= 2) |>
  arrange(desc(value_score)) |>
  mutate(
    Rank = row_number(),
    Salary = dollar(salary_m * 1e6, accuracy = 1),
    `Value Score` = round(value_score, 2),
    `Efficiency` = round(efficiency, 2),
    PPG = PTS,
    APG = AST,
    RPG = TRB
  ) |>
  select(Rank, Player, Team, Pos, Age, G, PPG, APG, RPG, 
         Salary, `Value Score`, Efficiency)

knitr::kable(elite_table, align = c("c","l","c","c","c","c",
                                     "c","c","c","r","c","c"))

```

Key Findings
=====================================

Row {data-height=800}
-------------------------------------

### Summary of Analysis

```{r}

findings <- tibble(
  `#` = 1:8,
  Finding = c(
    "Max contracts rarely justify their cost",
    "The NBA middle class has nearly vanished",
    "Point guards dominate value rankings",
    "Availability is undervalued in contracts",
    "Team payroll efficiency varies wildly",
    "Rotation players are the league's best value",
    "Efficiency drops as salary rises",
    "Dead money distorts team analysis"
  ),
  Detail = c(
    "Nearly all $40M+ players fall below the elite median value score. Only Luka Doncic and Nikola Jokic meaningfully exceed it among max earners.",
    "The $10-25M salary tier is virtually empty — NBA salaries are bimodal, clustering at the minimum and at max/near-max levels.",
    "PGs lead all positions in avg value score at 2.0, driven by contributions across points, assists and rebounds. Centers rank second.",
    "Value score scales by G/82. Players missing 20+ games on max contracts represent the clearest cases of salary inefficiency — availability is a skill.",
    "MEM and UTA lead the league in avg roster value score. LAL and LAC rank last, weighed down by aging max contracts and dead money.",
    "Top 15 value players all earn under $10M. Toumani Camara, Russell Westbrook and Collin Gillespie provide starter-level output on minimum deals.",
    "Across all positions and tiers, efficiency score shows a clear negative correlation with salary. Higher pay does not produce more efficient play.",
    "Players with contracts but zero games played (e.g. Anthony Davis, Trae Young on WAS) are included with $0 value — their salaries count against team payroll totals."
  )
)

knitr::kable(findings, align = c("c", "l", "l"))

```

---

**Data Source:** Basketball Reference 2025-26 NBA Player Stats (Per Game) and Player Contracts. Retrieved May 2026 from https://www.basketball-reference.com/leagues/NBA_2026_per_game.html and https://www.basketball-reference.com/contracts/players.html

**Methodology:** Value Score = (PTS + AST + TRB) / salary ($M) × (G/82). Efficiency Score = (PTS + TRB + AST + STL + BLK) / (missed FG + 0.44 × missed FT + TOV). Players earning less than $2M excluded from most analyses to remove two-way and Exhibit 10 contracts. Traded players use combined season totals. Players with active contracts but zero games played are included with $0 value score — their salaries count against team payroll totals. All salary figures reflect 2025-26 contractual values.